Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-15 talk/pepm2011/paper.tex
more suggestions by stephan

To understand the problems more directly, let us consider the simple
interpreter function \lstinline{f} that uses the object model (see the bottom of
Figure~\ref{fig:objmodel}).

The loop in \lstinline{f} iterates \lstinline{y} times, and computes something in the process.
Simply running this function is slow, because there are lots of virtual method
calls inside the loop, one for each \lstinline{is_positive} and even two for each
call to \lstinline{add}. These method calls need to check the type of the involved
objects repeatedly and redundantly. In addition, a lot of objects are created
when executing that loop, many of these objects are short-lived.
The actual computation that is performed by \lstinline{f} is simply a sequence of
float or integer additions.

@@

If the function is executed using the tracing JIT, with \lstinline{y} being a
\lstinline{BoxedInteger}, the produced trace looks like
Figure~\ref{fig:unopt-trace} (lines starting with the hash ``\#'' are comments).

The operations in the trace are shown indented to
correspond to the stack level of the function that contains the traced
operation. The trace is in single-assignment form, meaning that each variable is
assigned to exactly once. The arguments $p_0$ and $p_1$ of the loop correspond
to the live variables \lstinline{y} and \lstinline{res} in the original function.

@@

\begin{itemize}
    \item \lstinline{new} corresponds to object creation.
    \item \lstinline{get} correspond to attribute reads.
    \item \lstinline{set} correspond to attribute writes.
    \item \lstinline{guard_class} correspond to method calls and are followed by
    the trace of the called method.
    \item \lstinline{int_add} and \lstinline{int_gt} are integer addition and
    comparison (``greater than''), respectively.
\end{itemize}

The method calls in the trace are always preceded by a \lstinline{guard_class}
operation, to check that the class of the receiver is the same as the one that
was observed during tracing.\footnote{\lstinline{guard_class} performs a precise
class check, not checking for subclasses.} These guards make the trace specific
to the situation where \lstinline{y} is really a \lstinline{BoxedInteger}, it can
already be said to be specialized for \lstinline{BoxedIntegers}. When the trace is
turned into machine code and then executed with \lstinline{BoxedFloats}, the
first \lstinline{guard_class} instruction will fail and execution will continue
using the interpreter.

The trace shows the inefficiencies of \lstinline{f} clearly, if one looks at
the number of \lstinline{new}, \lstinline{set/get} and \lstinline{guard_class}
operations. The number of \lstinline{guard_class} operation is particularly
problematic, not only because of the time it takes to run them. All guards also
have additional information attached that makes it possible to return to the
interpreter, should the guard fail. This means that many guard operations also
lead to a memory problem.

@@

To understand the problems that this paper is trying to solve some more, we
first need to understand various cases of object lifetimes that can occur in a
tracing JIT compiler.

@@

Some of the operations within this trace are \lstinline{new} operations, which each
create a new instance of some class. These instances are used for a while, e.g.
by calling methods on them (which are inlined into the trace), reading and
writing their fields. Some of these instances \emph{escape}, which means that
they are stored in some globally accessible place or are passed into a
non-inlined function via a residual call.

@@

\begin{itemize}
    \item Category 1: Objects that live for a while, and are then just not
    used any more.

    \item Category 2: Objects that live for a while and then escape.

    \item Category 3: Objects that live for a while, survive across the jump to
    the beginning of the loop, and are then not used any more.

    \item Category 4: Objects that live for a while, survive across the jump,
    and then escape. To these we also count the objects that live across several
    jumps and then either escape or stop being used.
\end{itemize}

@@

The creation of objects in category 1 is removed by the optimization described
in Sections~\ref{sec:statics} and \ref{sec:formal}. Objects in the other
categories are partially optimized by this approach as well.

@@

In this case, the static heap afterwards would be
$$\{v^* \mapsto (T_1, w^*, v^*)\}$$.

>>>>

Let us now consider a simple interpreter function \lstinline{f} that uses the
object model (see the bottom of Figure~\ref{fig:objmodel}).
The loop in \lstinline{f} iterates \lstinline{y} times, and computes something in the process.
Simply running this function is slow, because there are lots of virtual method
calls inside the loop, one for each \lstinline{is_positive} and even two for each
call to \lstinline{add}. These method calls need to check the type of the involved
objects repeatedly and redundantly. In addition, a lot of objects are created
when executing that loop, many of these objects are short-lived.
The actual computation that is performed by \lstinline{f} is simply a sequence of
float or integer additions.

@@

If the function is executed using the tracing JIT, with \lstinline{y} being a
\lstinline{BoxedInteger}, the produced trace looks like the one of
Figure~\ref{fig:unopt-trace} (lines starting with a hash ``\#'' are comments).
The trace corresponds to one iteration of the while-loop in \lstinline{f}.

The operations in the trace are indented
corresponding to the stack level of the function that contains the traced
operation. The trace is in single-assignment form, meaning that each variable is
assigned a value exactly once. The arguments $p_0$ and $p_1$ of the loop correspond
to the live variables \lstinline{y} and \lstinline{res} in the original function.

@@

\begin{itemize}
    \item \lstinline{new} creates a new object.
    \item \lstinline{get} reads an attribute of an object.
    \item \lstinline{set} writes to an attribute of an object.
    \item \lstinline{guard_class} precedes an (inlined) method call and is
    followed by the trace of the called method.
    \item \lstinline{int_add} and \lstinline{int_gt} are integer addition and
    comparison (``greater than''), respectively.
\end{itemize}

The method calls in the trace are always preceded by a \lstinline{guard_class}
operation, to check that the class of the receiver is the same as the one that
was observed during tracing.\footnote{\lstinline{guard_class} performs a precise
class check, not checking for subclasses.} These guards make the trace specific
to the situation where \lstinline{y} is really a \lstinline{BoxedInteger}. When
the trace is turned into machine code and afterwards executed with
\lstinline{BoxedFloat}, the
first \lstinline{guard_class} instruction will fail and execution will continue
using the interpreter.

The trace shows the inefficiencies of \lstinline{f} clearly, if one looks at
the number of \lstinline{new}, \lstinline{set/get} and \lstinline{guard_class}
operations. The number of \lstinline{guard_class} operation is particularly
problematic, not only because of the time it takes to run them. All guards also
have additional information attached that makes it possible to return to the
interpreter, should the guard fail. This means that too many guard operations also
consume a lot of memory.

@@

To understand the problems that this paper is trying to solve in more detail, we
first need to understand various cases of object lifetimes that can occur in a
tracing JIT compiler.

@@

Some of the operations within this trace are \lstinline{new} operations, which each
create a new instance of some class. These instances are used for some time, \eg
by calling methods on them (which are inlined into the trace), reading and
writing their fields. Some of these instances \emph{escape}, which means that
they are stored in some globally accessible place or are passed into a
non-inlined function via a residual call.

@@

\begin{enumerate}
    \item Objects that live for some time, and are then just not
    used any more afterwards.

    \item Objects that live for some time and then escape.

    \item Objects that live for some time, survive across the jump to
    the beginning of the loop, and are then not used any more.

    \item Objects that live for some time, survive across the jump,
    and then escape. To these we also count the objects that live across several
    jumps and then either escape or stop being used.
\end{enumerate}

@@

The creation of objects in category 1 is removed by the optimization described
in Sections~\ref{sec:statics} and \ref{sec:formal}. Objects in the other
categories are partially optimized by this approach as well.\footnote{We also started to
work on optimizing objects in category 3, which will be the subject of a later
paper.}

@@

In this case, the static heap afterwards would be:
$$\{v^* \mapsto (T_1, w^*, v^*)\}$$
